SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-02-19 paper.tex
paper updates & first crack at parallel scaling

@@

In~\cref{F:intel_source}, the extent of this speedup is detailed; we note that the speedup displayed is normalized per-machine, such that the speedup of the wide-vectorized OpenCL code can be readily assessed.
On both machines, the unvectorized OpenCL code is \textasciitilde\SIrange{1.25}{1.5}{$\times$} faster than the baseline parallel OpenMP code.
Additionally, the wide-vectorized OpenCL code is \textasciitilde\SIrange{2.5}{3}{$\times$} and \textasciitilde\SIrange{3.5}{4}{$\times$} faster than the OpenMP code for the \sse/ and \avx/ machines respectively.
In contrast,~\cref{F:pocl_source} shows the runtime of deep, wide and unvectorized OpenCL codes on the POCL runtime, as compared to OpenMP parallelization.
No speedup is achieved for either vectorization type on either CPU; indeed the OpenMP is faster in both cases.
The reasons for this lack of vectorization are quite technical and somewhat outside the scope of this work, however briefly speaking vectorization might be achieved on POCL via a combination of relaxed math (``-cl-fast-relaxed-math''), use of explicit OpenCL vector types and newer versions of LLVM (specifically, potential improvments to the vectorization module).
This will be explored more in future work, but for the moment POCL is still quite useful as a validation tool.

@@

\Cref{f:source_permutate} shows the effect of changing data-ordering patterns as well as the \conp/ or \conv/ formulation on the performance of source term evaluation in \texttt{pyJac}.
In~\cref{F:source_conpvsconv}, we see that choice of the constant-pressure or constant-volume formulation has little to no effect on runtime for OpenMP as well as the shallow-vectorized\slash unvectorized Intel OpenCL codes on the \avx/ machine.
Generally speaking, the difference between the \conp/ and \conv/ formulations has little effect on performance regardless of CPU\slash GPU choice, or even for Jacobian evaluation which will be explored later.
In contrast,~\Cref{F:source_cvsf} shows large speedups of ``C''-ordered data over ``F''-ordered data on the \avx/ machine; the speedup presented is normalized per language, i.e., the roughly \SIrange{1.5}{2}{$\times$} speedup of the ``C''-ordered OpenMP implementation is relative to the ``F''-order OpenMP baseline while the OpenCL speedups are given with respect to the ``F''-ordered unvectorized OpenCL code.
Additionally, we note that the ``C''-ordered wide-vectorization in~\cref{F:source_cvsf} utilizes the vectorized-data ordering described in~\cref{S:data}.
In general, the data-ordering has a much more profound affect on the vectorized-OpenCL codes; this is due to the nature of SIMD-operations, which are much more efficient when operating on contiguous (in memory) data rather than incurring an expensive memory ``gather'' operation.
The ``C''-ordered unvectorized OpenCL code (which uses the default unvectorized row-major data ordering) is \textasciitilde\SIrange{1.5}{2.5}{$\times$} faster than the unvectorized ``F''-ordered code, and is in some cases almost as fast as the ``F''-ordered wide-vectorized OpenCL implementation.
However, the ``C''-ordered wide-vectorization achieves speedups of \textasciitilde\SIrange{4}{7}{$\times$} demonstrating the value of the vectorized data ordering for CPU execution.

\begin{figure}[htb]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_conpvsconv.pdf}
      \caption{The mean runtime per condition for each chemical model using both the \conp/ and \conv/ formulations on Intel OpenCL and OpenMP for \avx/.}
      \label{F:source_conpvsconv}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_cvsf.pdf}
      \caption{The speedup achieved by ``C'' ordering over ``F'' ordering for Intel OpenCL and OpenMP on the \avx/ CPU.  The speedup presented is normalized per-language (OpenMP and OpenCL) to better assess the effect of the data-ordering.}
      \label{F:source_cvsf}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_gpu_conpvsconv.pdf}
      \caption{The affect on performance of``C'' vs ``F''-ordering for the wide-vectorized NVIDIA OpenCL code on both GPUs.  The speedup presented is normalized per-GPU to better assess the effect of the data-ordering on each GPU.}
      \label{F:source_conpvsconv}
  \end{subfigure}
  \caption{The effect of the \conp/ and \conv/ formulations, as well as ``C'' and ``F'' data-ordering on source term evaluation performance in \texttt{pyJac}.
	   It is noted the wide-vectorized ``C'' ordered OpenCL case corresponds to the vectorized data ordering described in~\cref{S:data}}
  \label{f:source_permutate}

>>>>

\subsubsection{Source term evaluation}
\label{S:source_results}

@@

In~\cref{F:intel_source}, the extent of this speedup is detailed; we note that the speedup displayed is calculated per-machine, such that the speedup of the wide-vectorized OpenCL code can be readily assessed.
On both machines, the unvectorized OpenCL code is \textasciitilde\SIrange{1.25}{1.5}{$\times$} faster than the baseline parallel OpenMP code.
Additionally, the wide-vectorized OpenCL code is \textasciitilde\SIrange{2.5}{3}{$\times$} and \textasciitilde\SIrange{3.5}{4}{$\times$} faster than the OpenMP code for the \sse/ and \avx/ machines respectively.
In contrast,~\cref{F:pocl_source} shows the runtime of deep, wide and unvectorized OpenCL codes on the POCL runtime, as compared to OpenMP parallelization.
No speedup is achieved for either vectorization type on either CPU; indeed the OpenMP is faster in both cases.
The reasons for this lack of vectorization are quite technical and somewhat outside the scope of this work, however briefly speaking vectorization might be achieved on POCL via a combination of relaxed math (``-cl-fast-relaxed-math''), use of explicit OpenCL vector types and newer versions of LLVM (specifically, potential improvments to the vectorization module).
This will be explored more in future work, but for the moment POCL is still quite useful as a validation tool.

@@

\Cref{f:source_permutate} shows the effect of changing data-ordering patterns as well as the \conp/ or \conv/ formulation on the performance of source term evaluation in \texttt{pyJac}.
In~\cref{F:source_conpvsconv}, we see that choice of the constant-pressure or constant-volume formulation has little to no effect on runtime for OpenMP as well as the shallow-vectorized\slash unvectorized Intel OpenCL codes on the \avx/ machine.
Generally speaking, the difference between the \conp/ and \conv/ formulations has little effect on performance regardless of CPU\slash GPU choice, or even for Jacobian evaluation which will be explored later.

In contrast,~\Cref{F:source_cvsf} shows large speedups of ``C''-ordered data over ``F''-ordered data on the \avx/ machine; the speedup presented is calculated per language, i.e., the roughly \SIrange{1.5}{2}{$\times$} speedup of the ``C''-ordered OpenMP implementation is relative to the ``F''-order OpenMP baseline while the OpenCL speedups are given with respect to the ``F''-ordered unvectorized OpenCL code.
Additionally, we note that the ``C''-ordered wide-vectorization in~\cref{F:source_cvsf} utilizes the vectorized-data ordering described in~\cref{S:data}.
In general, the data-ordering has a much more profound affect on the vectorized-OpenCL codes; this is due to the nature of SIMD-operations, which are much more efficient when operating on contiguous (in memory) data rather than incurring an expensive memory ``gather'' operation.
The ``C''-ordered unvectorized OpenCL code (which uses the default unvectorized row-major data ordering) is \textasciitilde\SIrange{1.5}{2.5}{$\times$} faster than the unvectorized ``F''-ordered code, and is in some cases almost as fast as the ``F''-ordered wide-vectorized OpenCL implementation.
However, the ``C''-ordered wide-vectorization achieves speedups of \textasciitilde\SIrange{4}{7}{$\times$} demonstrating the value of the vectorized data ordering for CPU execution.

\Cref{F:source_gpu_cvsf} shows the effect of the ``C'' versus ``F''-ordering on source term evaluation performance on both GPUs.
The speedup reported is calculated per-GPU, such that the effect of data ordering can be seen on each individual GPU.
In general, the ``C''-ordered wide-vectorization is \textasciitilde\SI{1.05}{1.2}{$\times$} faster on the \newgpu/ GPU while the ``C'' and ``F''-ordered data are nearly equivalent for the \gpuold/ GPU, except for the isopentanol model where the ``F''-ordered data is \textasciitilde\SI{1.4}{$\times$} faster (though, we note that there was high variablity in this case). \todo{again, pending rerun on \gpuold/}
The ``C''-ordered speedup on the \gpunew/ GPU is counter to what one might expect, as typically speaking coalesced memory access in a wide-vectorization is easier to achieve if ``F''-ordering is used (see~\cref{S:data}).
However, the vectorized-data ordering here ensures that not only are memory accesses aligned to the vector width size (and thus, likely coalesced) but the improved data-locality; coupled with the larger cache sizes of the \gpunew/ GPU result in this modest speedup for ``C''-data ordering.

\begin{figure}[htb]
   \centering
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_conpvsconv.pdf}
      \caption{The mean runtime per condition for each chemical model using both the \conp/ and \conv/ formulations on Intel OpenCL and OpenMP for \avx/.}
      \label{F:source_conpvsconv}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_cvsf.pdf}
      \caption{The speedup achieved by ``C'' ordering over ``F'' ordering for Intel OpenCL and OpenMP on the \avx/ CPU.  The speedup presented is calculated per-language (OpenMP and OpenCL) to better assess the effect of the data-ordering.}
      \label{F:source_cvsf}
  \end{subfigure}
  \\
  \begin{subfigure}[t]{0.48\linewidth}
      \includegraphics[width=\textwidth]{source_gpu_cvsf.pdf}
      \caption{The affect on performance of ``C'' vs ``F''-ordering for the wide-vectorized NVIDIA OpenCL code on both GPUs with a vector width of 128.  The speedup presented is calculated per-GPU to better assess the effect of the data-ordering on each GPU.}
      \label{F:source_gpu_cvsf}
  \end{subfigure}
  \caption{The effect of the \conp/ and \conv/ formulations, as well as ``C'' and ``F'' data-ordering on source term evaluation performance in \texttt{pyJac}.
	   It is noted the wide-vectorized ``C'' ordered OpenCL case corresponds to the vectorized data ordering described in~\cref{S:data}}
  \label{f:source_permutate}
